The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-10 paper/ms.tex
Other small changes throughout

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the \textit{Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical position of $\sim15\,\pc$ for a thin disk
orbit or $\sim120\,\pc$ for a thick disk orbit is sufficient to induce a
$25\,\%$ systematic error in the vertical action $J_z$ computed under the
assumption of an axisymmetric potential. The induced error distribution is
non-Gaussian and bimodal, with neither of the modes peaking on the null value.
Furthermore, we show that the local midplane varies by $\sim100\,\pc$ at the
Solar circle in Milky Way-mass cosmological zoom-in simulations from the FIRE
project. From observations that the mean vertical velocity of the Milky Way
disk varies by $\sim5\textup{--}10\,\kms$ with radius, we estimate that the true midplane
variations are of order $\sim60\textup{--}170\,\pc$. Similar offsets are seen in three-dimensional dust maps. Thus, current
state-of-the-art action calculations --- which assume the global and local
midplanes are the same --- are likely to include a systematic vertical offset
depending on the volume considered. Variation in the local standard of rest
induces similar issues. The variation of the midplane must be taken into
account when performing dynamical modeling across the large regions of the
disk accessible to \textit{Gaia} and future missions.

@@

Before turning to numerical methods, we derive analytic expressions for the
systematic error in the actions induced from offsets in the position (e.g., the midplane or
Galactic center distance) or velocity (e.g., the local standard of rest) of the zero-point in the assumed coordinate system. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of the thin disk and a good description of the thick disk in
an axisymmetric potential, i.e., ignoring the influence of the Galactic bar and
spiral arms. We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle
\citep[e.g.,][]{2017MNRAS.465...76M}.

@@

We specifically consider \emph{azimuthal} variations in the midplane at the Solar circle, as defined by the stellar mass density. Since, to our knowledge, there are no direct empirical measurements of these variations in the Milky Way, we use example simulations from two classes of simulations to estimate the size of this effect.

One set are three zoom-in, cosmological-hydrodynamical simulations of Milky Way-mass galaxies from the FIRE collaboration, described briefly in \S \ref{ssec:cosmozoom}. These include stars, gas, and dark matter in a fully cosmological setting but are not tailored to specific properties of the Milky Way (such as the scale height or scale length, or the details of the accretion history). We use these simulations to span the range of possibilities for azimuthal midplane variations.

The other set of simulations are isolated N-body simulations of interactions between the Milky Way and a Sagittarius-like dwarf galaxy, described briefly in \S \ref{ssec:sag_sim}. These include dark matter and stars and are tailored to existing measurements of the structure of the Milky Way's disk and of the orbit and properties of the Sagittarius dwarf galaxy. Comparing the azimuthal midplane variations in the host galaxy of these simulations before and after the interaction with the Sagittarious-like object gives an idea of the effect of one minor merger whose properties are relatively well measured. Azimuthal variations of the mean vertical height of stars has been explicitly pointed out in a different simulation of a Sagittarius-like encounter by \citet{2013MNRAS.429..159G}.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from live N-body simulations of a
Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. The upper panels
show the midplane as a function of azimuth before the first encounter near the
Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$, with an encounter happening close to the Solar circle near $t=6.0\,\Gyr$. The fact that the $t=6.9\,\Gyr$ panel, which shows the midplane variation after some relaxation,
looks qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) is evidence that midplane variations are
generated, in part, by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

Figure~\ref{fig:midplane} shows the relative $z$ location of the inferred
midplane an observer would determine as a function of azimuth for
each galaxy, using their local ``solar neighborhood'' (the cylinder defined
above). We consider $50$ equally spaced bins in azimuth, a sufficient number
such that each cylinder shares no stars with its neighboring cylinder. The
$1\,\sigma$ error from the bootstrap procedure is shown as the dashed-line
error bars. The middle $90\%$ of midplane values across the Solar circle spans
$185$, $162$, $84\,\pc$ for \mi{}, \mf{}, and \mm{}. In two of the three cases
the midplane therefore varies by more than $\pm 100\,\pc$ depending on the
azimuth along the Solar circle; in the third (\mm{}, which has the thinnest
``thin disk'' of stars, but the largest stellar mass) the variation is closer
to $\pm 50\,\pc$.

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for four succcessive timesteps of the live N-body simulation of a
Sagittarius encounter \citep{2018MNRAS.481..286L}. Again we have subtracted a
best fit curve of the form $\text{A} \sin{\left(\phi + \text{B}\right)} +
\text{C}$, with the values for $\text{A}$ being $9.5$, $2.5$, $-212$, and
$-394\,\pc$, for $\text{B}$ being $0.0013$, $0.0007$, $-0.63$, and
$-1.00\,\text{rad}$ and $\text{C}$ being $2.6$, $-7.8$, $-65$, and $-53\,\pc$
chronologically for each of the four timesteps shown. The middle $90\%$ of midplane values across the Solar circle spans $49$, $62$, $137$, and $119\,\pc$, for each timestep.

@@

\subsubsection{Local Standard of Rest}
Finally, one must define the LSR, or mean velocity of stars near the Sun
relative to the Galactic center (which is defined to have zero velocity), and
the velocity of the Sun relative to the LSR. The radial ($U_{\odot}$) and
vertical ($W_{\odot}$) components are computed by taking the mean motions of
different stellar groups \citep[e.g.,][]{2012MNRAS.427..274S}. The azimuthal
component ($V_{\odot}$) is more difficult to measure, but can be modeled using
the asymmetric drift relation \citep{2008gady.book.....B}. The values of the
components of the LSR are usually taken from \citet{2010MNRAS.403.1829S}. Their uncertainties should also lead to systematic errors in the actions, as given in Eqs. \ref{eq:induced_Jphi}--\ref{eq:induced_Jz}. For example, the
value of the circular velocity is taken to be $\sim 220\kms$
\citep[e.g.,][]{2012ApJ...759..131B} with roughly $10\%$ uncertainty. We expect this to translate to at least a $10\%$ systematic error in $J_{\phi}$.

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the Milky Way with Sagittarius. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim185$, $162$, $84\,\pc$ (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim60$ and $\sim120\,\pc$ before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the Milky Way suggests that these values are consistent with expectations for midplane variations in the Milky Way itself, which have not yet been measured as a function of azimuth.

@@

\item Inaccuracies in the parameters of the currently adopted
Galactocentric coordinate system are likely important for some
applications. In particular, it is imperative to test the assumption that the
dynamical Galactic center is colocated with \sgra{}. We discuss how to
do this in \S \ref{ssec:coord_off}.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/lsr.pdf}
\end{center}
\caption{The LSR as a function of azimuth at the
fiducial Solar circle ($R_0 = 8.2\,\kpc$). No best-fit subtraction is
performed here as we did in the case of the midplane
(Section~\ref{ssec:local_midplane}). Variations in $v_z$ are on the order of
$\sim5\textup{--}10\,\kms$.}
\label{fig:lsr_variations}
\end{figure*}

>>>>

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the \textit{Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical position of $\sim15\,\pc$ for a thin disk
orbit or $\sim120\,\pc$ for a thick disk orbit is sufficient to induce a
$25\,\%$ systematic error in the vertical action $J_z$ computed under the
assumption of an axisymmetric potential. The induced error distribution is
non-Gaussian and bimodal, with neither of the modes peaking on the null value.
Furthermore, we show that the local midplane varies by $\sim100\,\pc$ at the
solar circle in Milky Way-mass cosmological zoom-in simulations from the FIRE
project. From observations that the mean vertical velocity of the Milky Way
disk varies by $\sim5\textup{--}10\,\kms$ with radius, we estimate that the true midplane
variations are of order $\sim60\textup{--}170\,\pc$. Similar offsets are seen in three-dimensional dust maps. Thus, current
state-of-the-art action calculations --- which assume the global and local
midplanes are the same --- are likely to include a systematic vertical offset
depending on the volume considered. Variation in the local standard of rest
induces similar issues. The variation of the midplane must be taken into
account when performing dynamical modeling across the large regions of the
disk accessible to \textit{Gaia} and future missions.

@@

Before turning to numerical methods, we derive analytic expressions for the
systematic error in the actions induced from offsets in the position (e.g., the midplane or
Galactic center distance) or velocity (e.g., the local standard of rest) of the zero-point in the assumed coordinate system. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of the thin disk and a good description of the thick disk in
an axisymmetric potential, i.e., ignoring the influence of the Galactic bar and
spiral arms. We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the solar circle
\citep[e.g.,][]{2017MNRAS.465...76M}.

@@

We specifically consider \emph{azimuthal} variations in the midplane at the solar circle, as defined by the stellar mass density. Since, to our knowledge, there are no direct empirical measurements of these variations in the Milky Way, we use example simulations from two classes of simulations to estimate the size of this effect.

One set are three zoom-in, cosmological-hydrodynamical simulations of Milky Way-mass galaxies from the FIRE collaboration, described briefly in Section~\ref{ssec:cosmozoom}. These include stars, gas, and dark matter in a fully cosmological setting but are not tailored to specific properties of the Milky Way (such as the scale height or scale length, or the details of the accretion history). We use these simulations to span the range of possibilities for azimuthal midplane variations.

The other set of simulations are isolated N-body simulations of interactions between the Milky Way and a Sagittarius-like dwarf galaxy, described briefly in Section~\ref{ssec:sag_sim}. These include dark matter and stars and are tailored to existing measurements of the structure of the Milky Way's disk and of the orbit and properties of the Sagittarius dwarf galaxy. Comparing the azimuthal midplane variations in the host galaxy of these simulations before and after the interaction with the Sagittarious-like object gives an idea of the effect of one minor merger whose properties are relatively well measured. Azimuthal variations of the mean vertical height of stars has been explicitly pointed out in a different simulation of a Sagittarius-like encounter by \citet{2013MNRAS.429..159G}.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from live N-body simulations of a
Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. The upper panels
show the midplane as a function of azimuth before the first encounter near the
solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$, with an encounter happening close to the solar circle near $t=6.0\,\Gyr$. The fact that the $t=6.9\,\Gyr$ panel, which shows the midplane variation after some relaxation,
looks qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) is evidence that midplane variations are
generated, in part, by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

Figure~\ref{fig:midplane} shows the relative $z$ location of the inferred
midplane an observer would determine as a function of azimuth for
each galaxy, using their local ``solar neighborhood'' (the cylinder defined
above). We consider $50$ equally spaced bins in azimuth, a sufficient number
such that each cylinder shares no stars with its neighboring cylinder. The
$1\,\sigma$ error from the bootstrap procedure is shown as the dashed-line
error bars. The middle $90\%$ of midplane values across the solar circle spans
$185$, $162$, $84\,\pc$ for \mi{}, \mf{}, and \mm{}. In two of the three cases
the midplane therefore varies by more than $\pm 100\,\pc$ depending on the
azimuth along the solar circle; in the third (\mm{}, which has the thinnest
``thin disk'' of stars, but the largest stellar mass) the variation is closer
to $\pm 50\,\pc$.

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for four succcessive timesteps of the live N-body simulation of a
Sagittarius encounter \citep{2018MNRAS.481..286L}. Again we have subtracted a
best fit curve of the form $\text{A} \sin{\left(\phi + \text{B}\right)} +
\text{C}$, with the values for $\text{A}$ being $9.5$, $2.5$, $-212$, and
$-394\,\pc$, for $\text{B}$ being $0.0013$, $0.0007$, $-0.63$, and
$-1.00\,\text{rad}$ and $\text{C}$ being $2.6$, $-7.8$, $-65$, and $-53\,\pc$
chronologically for each of the four timesteps shown. The middle $90\%$ of midplane values across the solar circle spans $49$, $62$, $137$, and $119\,\pc$, for each timestep.

@@

\subsubsection{Local Standard of Rest}
Finally, one must define the LSR, or mean velocity of stars near the Sun
relative to the Galactic center (which is defined to have zero velocity), and
the velocity of the Sun relative to the LSR. The radial ($U_{\odot}$) and
vertical ($W_{\odot}$) components are computed by taking the mean motions of
different stellar groups \citep[e.g.,][]{2012MNRAS.427..274S}. The azimuthal
component ($V_{\odot}$) is more difficult to measure, but can be modeled using
the asymmetric drift relation \citep{2008gady.book.....B}. The values of the
components of the LSR are usually taken from \citet{2010MNRAS.403.1829S}. Their uncertainties should also lead to systematic errors in the actions, as given in Equations~\eqref{eq:induced_Jphi}--\eqref{eq:induced_Jz}. For example, the
value of the circular velocity is taken to be $\sim 220\kms$
\citep[e.g.,][]{2012ApJ...759..131B} with roughly $10\%$ uncertainty. We expect this to translate to at least a $10\%$ systematic error in $J_{\phi}$.

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the Milky Way with Sagittarius. We found
that the midplane varies as a function of azimuth at the solar circle by
$\sim185$, $162$, $84\,\pc$ (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim60$ and $\sim120\,\pc$ before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the Milky Way suggests that these values are consistent with expectations for midplane variations in the Milky Way itself, which have not yet been measured as a function of azimuth.

@@

\item Inaccuracies in the parameters of the currently adopted
Galactocentric coordinate system are likely important for some
applications. In particular, it is imperative to test the assumption that the
dynamical Galactic center is colocated with \sgra{}. We discuss how to
do this in Section~\ref{ssec:coord_off}.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/lsr.pdf}
\end{center}
\caption{The LSR as a function of azimuth at the
fiducial solar circle ($R_0 = 8.2\,\kpc$). No best-fit subtraction is
performed here as we did in the case of the midplane
(Section~\ref{ssec:local_midplane}). Variations in $v_z$ are on the order of
$\sim5\textup{--}10\,\kms$.}
\label{fig:lsr_variations}
\end{figure*}
